---
title: "Differences in racial burdens of COVID deaths"
author: "Adeline Lo and Héctor Pifarré i Arolas"
date: "This version: April 2021"
output: 
  html_document: 
    code_folding: hide
    pandoc_args: ["--lua-filter=color-text.lua"]
    toc: true
    toc_float: true
    number_sections: true
    theme: united
  pdf_document: 
    pandoc_args: ["--lua-filter=color-text.lua"]
    keep_tex: true
header-includes:
  - \preauthor{\centering\large}
  - \predate{\centering\normalsize}
  - \pretitle{\centering\Large\textbf}
  - \usepackage{setspace}\onehalfspacing
  - \usepackage{xcolor}
  - \usepackage[shortlabels]{enumitem}
  - \usepackage{pgf,tikz, mathabx}
  - \usetikzlibrary{positioning}
editor_options:
  chunk_output_type: console
---
<style>

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}

</style>

```{cat, engine.opts = list(file = "color-text.lua")}
Span = function(span)
  color = span.attributes['color']
  -- if no color attribute, return unchange
  if color == nil then return span end
  
  -- tranform to <span style="color: red;"></span>
  if FORMAT:match 'html' then
    -- remove color attributes
    span.attributes['color'] = nil
    -- use style attribute instead
    span.attributes['style'] = 'color: ' .. color .. ';'
    -- return full span element
    return span
  elseif FORMAT:match 'latex' then
    -- remove color attributes
    span.attributes['color'] = nil
    -- encapsulate in latex code
    table.insert(
      span.content, 1,
      pandoc.RawInline('latex', '\\textcolor{'..color..'}{')
    )
    table.insert(
      span.content,
      pandoc.RawInline('latex', '}')
    )
    -- returns only span content
    return span.content
  else
    -- for other format return unchanged
    return span
  end
end
```

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
# rm(list=ls())
library(forcats)
library(devtools)
library(cspp)
library(data.table)
library(rapportools)
library(tidyverse)
library(lubridate)
#library(plyr)
library('tidyr')
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(hrbrthemes)
library(viridis)
#library(xlsx)
library(glue)
library(dplyr)
library(openxlsx)
library(lubridate)
library(anytime)
```

# COVID-Deaths by Race, Age, Time

Access CDC *AH Monthly provisional counts of deaths by age group and race/ethnicity for select causes of death* (https://data.cdc.gov/NCHS/AH-Monthly-provisional-counts-of-deaths-by-age-gro/r5pw-bk5t) 

* Only U.S. aggregated
* Can't find dataset on CDC that is broken down by all of the following: State/Time/Race/Age 


```{r call-in-data}
covidraceage<-read.csv("data/AH_Monthly_provisional_counts_of_deaths_by_age_group_and_race_ethnicity_for_select_causes_of_death.csv",stringsAsFactors = FALSE)

#table(covidraceage$Age.Group)10 age groups. 


#rename some variables for ease:
covidraceage<-covidraceage %>% 
  dplyr::select(-starts_with("flag_")) %>%
  dplyr::rename(
    StartDate = Start.Date,
    EndDate = End.Date,
    Race = Race.and.Hispanic.Origin.Group,
    Age = Age.Group,
    COVID_MultCauseDeath = COVID.19..Multiple.Cause.of.Death.,
    COVID_CauseDeath = COVID.19..Underlying.Cause.of.Death.
    )

covidraceage$StartDate<-mdy(covidraceage$StartDate)
covidraceage$EndDate<-mdy(covidraceage$EndDate)

covidraceage %>%
  as_tibble()


# Careful with this definition of covid deaths! Let's just use covid deaths without multicause death, as it is the common definition

#covidraceage$COVID_PartofDeath<-covidraceage$COVID_CauseDeath+covidraceage$COVID_MultCauseDeath
covidraceage$COVID_PartofDeath<-covidraceage$COVID_CauseDeath

#sum for each race/age/month/year the covid deaths

age_ranges<-c("0-4 years","5-14 years","15-24 years","25-34 years","35-44 years","45-54 years","55-64 years","65-74 years"
            ,"75-84 years","85 years and over")
races<-unique(covidraceage$Race)
DF<-vector("list",length=length(races))
#we want to end up with list of race dataframes, each race dataframe is 10 age grows rows, columns are Month-Year, elements are covid deaths.

# informationholder for time: matches 2020-01 --> 2021-02 14 months --> 1-14
# sum(covidraceage$COVID_CauseDeath)

covidraceage_new <- covidraceage %>% 
  filter(Year == "2020" |Year == "2021")

covidraceage_new$Month[covidraceage_new$Year == "2021" & covidraceage_new$Month == "1"] <- 13
covidraceage_new$Month[covidraceage_new$Year == "2021" & covidraceage_new$Month == "2"] <- 14

covidraceage_white <- covidraceage_new %>%
  filter(Race == "Non-Hispanic White") %>%
  dplyr::select(Age,Year,Month,COVID_PartofDeath) 

covidraceage_black <- covidraceage_new %>%
  filter(Race == "Non-Hispanic Black") %>%
  dplyr::select(Age,Year,Month,COVID_PartofDeath)

covidraceage_hispanic <- covidraceage_new %>%
  filter(Race == "Hispanic") %>%
  dplyr::select(Age,Year,Month,COVID_PartofDeath)

covidraceage_asian <- covidraceage_new %>%
  filter(Race == "Non-Hispanic Asian") %>%
  dplyr::select(Age,Year,Month,COVID_PartofDeath)

saveRDS(covidraceage_white,file="../Data/covidraceage_white_unweighted.rds")
saveRDS(covidraceage_black,file="../Data/covidraceage_black_unweighted.rds")
saveRDS(covidraceage_hispanic,file="../Data/covidraceage_hispanic_unweighted.rds")
saveRDS(covidraceage_asian,file="../Data/covidraceage_asian_unweighted.rds")
#table(covidraceage$Race) "Hispanic","Non-Hispanic Asian","Non-Hispanic Black","Non-Hispanic White" (leave out: "Other"  "Non-Hispanic American Indian or Alaska Native")



## Finish creation of 4 dataframes (one for each race), each dataframe is 10 x 14 (age groups x months from 2020-01 -> 2021-02), each cell is COVID_PartofDeath count.
```


# Get standard populations for U.S. 
* from https://seer.cancer.gov/stdpopulations/stdpop.singleages.html and clean it.

```{r get-usstdpop, eval=F}
library(htmltab)
url <- "https://seer.cancer.gov/stdpopulations/stdpop.singleages.html"
usstdpop <- htmltab(doc = url)
usstdpop %>%
  as_tibble()
# saveRDS(usstdpop,file="../Data/StdPop/usstdpop.rds")
```

```{r clean-usstdpop-popagerace, echo=FALSE, message=FALSE, warning=FALSE}
# usstdpop<-readRDS(file="../Data/StdPop/usstdpop.rds")


#use first two cols, `Age` & `2000 U.S. Standard Population (Census P25-1130) >> Single Ages to 99`

usstdpop$USStdPop99<-as.numeric(gsub(pattern=",",replacement="",x=usstdpop$`2000 U.S. Standard Population (Census P25-1130) >> Single Ages to 99`) ) #std pop variable for each year-increment of age
usstdpop$Age_no<-gsub(pattern=" years",replacement="",x=usstdpop$Age)

StdPop<-data.frame(Age=usstdpop$Age_no[which(usstdpop$Age_no!="Total")]
           ,stdpop=usstdpop$USStdPop99[which(usstdpop$Age_no!="Total")])
StdPop$Age[which(StdPop$Age=="100+")]<-100
StdPop$Age<-as.numeric(StdPop$Age)
saveRDS(StdPop,file="../Data/StdPop/StdPop.rds")
#sum up by age categories that the CDC has using #table(covidraceage$Age.Group)

Group_AgeStdPop<-data.frame(Age=c("0-4 years","5-14 years","15-24 years","25-34 years","35-44 years","45-54 years","55-64 years","65-74 years"
            ,"75-84 years","85 years and over"),
Grouped_StdPop=c(sum(StdPop$stdpop[StdPop$Age<=4]),sum(StdPop$stdpop[StdPop$Age>4&StdPop$Age<=14])
                 ,sum(StdPop$stdpop[StdPop$Age>14&StdPop$Age<=24]),sum(StdPop$stdpop[StdPop$Age>24&StdPop$Age<=34])
                 ,sum(StdPop$stdpop[StdPop$Age>34&StdPop$Age<=44]),sum(StdPop$stdpop[StdPop$Age>44&StdPop$Age<=54])
                 ,sum(StdPop$stdpop[StdPop$Age>54&StdPop$Age<=64]),sum(StdPop$stdpop[StdPop$Age>64&StdPop$Age<=74])
                 ,sum(StdPop$stdpop[StdPop$Age>74&StdPop$Age<=84]),sum(StdPop$stdpop[StdPop$Age>84]))
)
Group_AgeStdPop$AgeStdPop_Weight<-Group_AgeStdPop$Grouped_StdPop/sum(Group_AgeStdPop$Grouped_StdPop) #weight sums to 1

popraceage<-read.csv("pnas_alternative/pop-age-race_frompnas.csv")

Group_AgeStdPop$Hispanic_pop_age<-popraceage$Hispanic
Group_AgeStdPop$White_pop_age<-popraceage$non.Hispanic.White
Group_AgeStdPop$Black_pop_age<-popraceage$non.Hispanic.Black
```

$$R_{g,t} = \sum_i  (r_{i,t,g} \times w_i)$$

where 

$$r_{i,t,g} = \frac{\text{COVID deaths in this age group}_{t,g}}{\text{Population in this age group}_{g}}$$
and 
$$w_i = \frac{\text{Population}_i}{\sum_i \text{Population}_i}$$

We're going to calculate $R_{g,t}$ for race groups $g$ Hispanic, Non-Hispanic White, and Black (since we have age range population for these three groups but not Asian). `AgeStdPop_weight` in the above is equivalent to $w_i$. For $r_{i,t,g}$ we get these from the covid counts for each race group $g$, age group $i$, time $t$, then divide by the population in the age group $i$ for the race group $g$. This latter part is from the pnas file `pnas.2014746118.sd01 (1)` sheet 10, "Population Counts", used in the article https://www.pnas.org/content/118/5/e2014746118. We've pulled the first table of population counts by race into a separate csv file for ease of import, called `pop-age-race_frompnas.csv`, and with the 5to9 and 10to14 age groups summed up so they match our CDC covid-death counts data age groups perfectly. Covid death counts for $i$ for race $g$ divided by `Population in this age group_{g}` and multiplied by 100,000 is $r_{i,t,g}$ in the above (called `Group_AgeStdPop$Race_pop_age` for the appropriate `Race` -- Hispanic, White, Black).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
covidraceage_white <- readRDS(file = "../Data/covidraceage_white_unweighted.rds")
covidraceage_black<- readRDS(file = "../Data/covidraceage_black_unweighted.rds")
covidraceage_hispanic<- readRDS(file = "../Data/covidraceage_hispanic_unweighted.rds")
covidraceage_asian<- readRDS(file = "../Data/covidraceage_asian_unweighted.rds")

covidwhite_weighted <- right_join(covidraceage_white,Group_AgeStdPop) %>%
  mutate(riwi_white = (COVID_PartofDeath/White_pop_age)*100000*AgeStdPop_Weight)%>% #this is r_i x w_i
  dplyr::select(-Grouped_StdPop,-AgeStdPop_Weight) %>%
  rename(COVID_PartofDeath_white = COVID_PartofDeath)

covidblack_weighted <- right_join(covidraceage_black,Group_AgeStdPop) %>%
  mutate(riwi_black = (COVID_PartofDeath/Black_pop_age)*100000*AgeStdPop_Weight)%>% #this is r_i x w_i
  dplyr::select(-Grouped_StdPop,-AgeStdPop_Weight) %>%
  rename(COVID_PartofDeath_black = COVID_PartofDeath)

covidhispanic_weighted <- right_join(covidraceage_hispanic,Group_AgeStdPop) %>%
  mutate(riwi_hispanic = (COVID_PartofDeath/Hispanic_pop_age)*100000*AgeStdPop_Weight)%>% #this is r_i x w_i
  dplyr::select(-Grouped_StdPop,-AgeStdPop_Weight) %>%
  rename(COVID_PartofDeath_hispanic = COVID_PartofDeath)
  

covid_race_weighted <- list(covidwhite_weighted, covidblack_weighted, covidhispanic_weighted) %>% reduce(right_join, by = c("Age","Year","Month"))

covid_race_weighted2 <- covid_race_weighted %>%
  #group_by(Age,Year,Month) %>%
  group_by(Year,Month) %>%
  dplyr::summarise(White = sum(riwi_white),
            Black = sum(riwi_black),
            Hispanic = sum(riwi_hispanic)) %>%
  pivot_longer(c(White,Black,Hispanic),
               names_to = "Race",
               values_to = "COVID",
  )

ggplot(data = covid_race_weighted2) + 
  geom_line (aes(x = Month, y = COVID, colour = Race), size=1.25) +
  scale_x_continuous(breaks = 1:14) + theme_bw() +  scale_color_viridis(option="D",alpha=1,discrete=TRUE) +
  ylab("COVID Death Rate per 100,000 (Age Standardized)") + xlab("Months from January 2020")

#Figures/us_race_time_coviddeathrate.pdf

# Adding overall deaths, simply sum the counts for all the races
# Reading in overall file
# covidraceage_new earlier
# Summary by month
counts_summary<-covidraceage_new %>%
             group_by(Year,Month) %>%   
  #simple counts by race and week
  summarize(Death_counts=COVID_PartofDeath)
# Replacing NAs with 0s
counts_summary[is.na(counts_summary)] <- 0

# Counts data frame for overall
counts_overall<-data.frame(Year=c(rep(2020,12),rep(2021,2)),
                           Month=seq(1:14))
# US population data
total_us<-328239523
# Now summing up across races
for (i in 1:length(unique(counts_overall$Month))){
  for (j in 2020:2021){
  counts_overall$Death_counts[which(counts_overall$Month==i&counts_overall$Year==j)]<-
    sum(counts_summary$Death_counts[which(counts_summary$Month==i&counts_summary$Year==j)])
  }                                                                                                                        
}
# Checking totals  
#sum(counts_overall$Death_counts)
# Plots of overall death counts
ggplot(counts_overall)+
  geom_line(aes(x= Month,y = Death_counts), color = "black", size=1.25)+
  scale_x_continuous(breaks = 1:14) + theme_bw() +  scale_color_viridis(option="D",alpha=1,discrete=TRUE) +
  ylab("COVID Death Counts") + xlab("Months from January 2020")




```


