---
title: "Untitled"
author: "JIAYI HONG"
date: "5/28/2021"
output: pdf_document
---


```{r libraries, echo=FWISE, message=FWISE, warning=FWISE}
rm(list=ls())
library(forcats)
library(devtools)
library(cspp)
library(data.table)
library(rapportools)
#library(tidyverse)
library(lubridate)
library(plyr)
library('tidyr')
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(hrbrthemes)
library(viridis)
#library(xlsx)
library(glue)
library(dplyr)
library(openxlsx)
library(lubridate)
library(anytime)
```

```{r}
population_WI = read.delim("data/Single-Race Population Estimates 2010-2019 by State and Single-Year Age (21).txt", header = T, sep = "\t")


population_WI = population_WI %>%
  filter(States == "Wisconsin")
population_WI
unique(population_WI$Five.Year.Age.Groups)
unique(population_WI$Race)
unique(population_WI$Ethnicity)
```

```{r}
population_WI_new = population_WI

# combine age groups, and the new age group names are exactly the same as those of the death_?? data below
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "< 1 year")] = "0-14 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "1-4 years")] = "0-14 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "5-9 years")] = "0-14 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "10-14 years")] = "0-14 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "15-19 years")] = "15-29 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "20-24 years")] = "15-29 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "25-29 years")] = "15-29 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "30-34 years")] = "30-49 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "35-39 years")] = "30-49 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "40-44 years")] = "30-49 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "45-49 years")] = "30-49 Years"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "50-54 years")] = "50-64 Years" 
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "55-59 years")] = "50-64 Years" 
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "60-64 years ")] = "50-64 Years" 
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "65-69 years")] = "65+"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "70-74 years")] = "65+"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "75-79 years")] = "65+"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "80-84 years")] = "65+"
population_WI_new$Five.Year.Age.Groups[which(population_WI_new$Five.Year.Age.Groups == "85+ years")] = "65+"

colnames(population_WI_new)[4] = "age_groups" # Five.Year.Age.Groups => age_groups
#colnames(population_WI_new)

population_WI_new
```

```{r}
# age groups for all races (>3)
population_WI_all_races_sum = population_WI_new %>%
  group_by(age_groups) %>%
  dplyr::summarise(population = sum(Population))

population_WI_all_races_sum

# create racial groups and summarise by age
population_WI_white_sum = population_WI_new %>%
  filter(Race == "White" & Ethnicity == "Not Hispanic or Latino") %>%
  select(age_groups, Population) %>%
  group_by(age_groups) %>%
  dplyr::summarise(population = sum(Population))

population_WI_black_sum = population_WI_new %>%
  filter(Race == "Black or African American" & Ethnicity == "Not Hispanic or Latino") %>%
  select(age_groups, Population) %>%
  group_by(age_groups) %>%
  dplyr::summarise(population = sum(Population))

population_WI_hispanic_sum = population_WI_new %>%
  filter(Ethnicity == "Hispanic or Latino") %>%
  select(age_groups, Population) %>%
  group_by(age_groups) %>%
  dplyr::summarise(population = sum(Population))
```


```{r}
# overall death data
death = read.csv("data/AH_Quarterly_Excess_Deaths_by_State__Sex__Age__and_Race.csv")
table(death$AgeGroup)
death
```



```{r}
# adjust age groups
death$AgeGroup[which(death$AgeGroup == "50-54 Years")] = "50-64 Years" 
death$AgeGroup[which(death$AgeGroup == "50-64 Years")] = "50-64 Years" 

table(death$AgeGroup)
```



```{r}
death_new = death %>%
  filter(Year == "2020" |Year == "2021")
death_new$Quarter[death_new$YearQuarter == "2021, Quarter 1"] = "Quarter 5"
death_new$Quarter[death_new$YearQuarter == "2021, Quarter 2"] = "Quarter 6"

# by state
death_WI = death_new %>%
  filter(State == "WI" & AgeGroup != "all Ages" & AgeGroup != "Not stated")
  
# death in 3 racial groups, combined version
death_WI_3races_sum = death_WI %>%
  filter(RaceEthnicity == "Non-Hispanic White" | RaceEthnicity == "Non-Hispanic Black" | RaceEthnicity == "Hispanic") %>%
  group_by(AgeGroup, Quarter, RaceEthnicity) %>%
  dplyr::summarise(Deaths_weighted = sum(Deaths..weighted.))
death_WI_3races_sum
```

```{r, include=FalSE}
# death in racial groups, separated version
death_WI_white_sum = death_WI %>%
  filter(RaceEthnicity == "Non-Hispanic White") %>%
  select(AgeGroup,Quarter,Deaths..weighted.) %>%
  group_by(AgeGroup, Quarter) %>%
  dplyr::summarise(Deaths_weighted = sum(Deaths..weighted.))

death_WI_black_sum = death_WI %>%
  filter(RaceEthnicity == "Non-Hispanic Black") %>%
  dplyr::select(AgeGroup,Quarter,Deaths..weighted.) %>%
  group_by(AgeGroup, Quarter) %>%
  dplyr::summarise(Deaths_weighted = sum(Deaths..weighted.))

death_WI_hispanic_sum = death_WI %>%
  filter(RaceEthnicity == "Hispanic") %>%
  dplyr::select(AgeGroup,Quarter,Deaths..weighted.) %>%
  group_by(AgeGroup, Quarter) %>%
  dplyr::summarise(Deaths_weighted = sum(Deaths..weighted.))

death_WI_black_sum

# death_asian = death_new %>%
#   filter(RaceEthnicity == "Non-Hispanic Asian") %>%
#   dplyr::select(AgeGroup,Year,Quarter,Deaths..weighted.)
```

$$
R_{g,t} = \sum_i  (r_{i,t,g} \times w_i)
$$

where 

$$r_{i,t,g} = \cfrac{\text{COVID deaths in this age group}_{t,g}}{\text{Population in this age group}_{g}}$$
and 
$$w_i = \cfrac{\text{Population}_i}{\sum_i \text{Population}_i}$$

We're going to cWIculate $$R_{g,t}$$ for race groups $$g$$ Hispanic, Non-Hispanic White, and Black (since we have age range population for these three groups but not Asian). 


`AgeStdPop_weight` in the above is equivWIent to $w_i$. 


For $r_{i,t,g}$ we get these from the covid counts for each race group $g$, age group $i$, time $t$, then divide by the population in the age group $i$ for the race group $g$. This latter part is from the pnas file `pnas.2014746118.sd01 (1)` sheet 10, "Population Counts", used in the article https://www.pnas.org/content/118/5/e2014746118. We've pulled the first table of population counts by race into a separate csv file for ease of import, called `pop-age-race_frompnas.csv`, and with the 5to9 and 10to14 age groups summed up so they match our CDC covid-death counts data age groups perfectly. Covid death counts for $i$ for race $g$ divided by `Population in this age group_{g}` and multiplied by 100,000 is $r_{i,t,g}$ in the above (called `Group_AgeStdPop$Race_pop_age` for the appropriate `Race` -- Hispanic, White, Black).

```{r}

population_WI_all_races_sum



population_WI_all_races_sum # total population, including all races
population_WI_white_sum
population_WI_black_sum
population_WI_hispanic_sum
```

```{r}
population_WI_new
population_WI_total = sum(population_WI_new$Population)

# population_WI_total = sum(population_WI_black_sum$population)+sum(population_WI_white_sum$population)+sum(population_WI_hispanic_sum$population)

#population_WI_total
```

$$
w_i = \cfrac{\text{Population}_i}{\sum_i \text{Population}_i}
$$

```{r}
w_1 = population_WI_all_races_sum$population[1]/ population_WI_total
w_2 = population_WI_all_races_sum$population[2] / population_WI_total
w_3 = population_WI_all_races_sum$population[3]	 / population_WI_total
w_4 = population_WI_all_races_sum$population[4] / population_WI_total
w_5 = population_WI_all_races_sum$population[5] / population_WI_total

w_i = c(w_1, w_2, w_3, w_4, w_5)
age_ranges = c("0-14 Years","15-29 Years","30-49 Years","50-64 Years","65+")
names(w_i) = age_ranges
w_i
```



```{r}
death_WI_3races_sum
# death_WI_white_sum
# death_WI_black_sum
# death_WI_hispanic_sum
```

$$
r_{i,t,g} = \cfrac{\text{COVID deaths in this age group}_{t,g}}{\text{Population in this age group}_{g}}
$$

```{r}
# look for population with specific age and race

########## change this function for different states

r_igt_lower_func = function(age, race) {
  if (race == "Non-Hispanic White") {df = population_WI_white_sum}
  if (race == "Hispanic") {df = population_WI_hispanic_sum}
  if (race == "Non-Hispanic Black") {df = population_WI_black_sum}
  
  if (age == "0-14 Years") {lower = df$population[1]}
  if (age == "15-29 Years") {lower = df$population[2]}
  if (age == "30-49 Years") {lower = df$population[3]}
  if (age == "50-64 Years") {lower = df$population[4]}
  if (age == "65+") {lower = df$population[5]}
  
  return(lower)
}
```

```{r}
death_WI_3races_sum_rigt = death_WI_3races_sum %>%
  mutate(r_igt = 0)

for (i in c(1:nrow(death_WI_3races_sum_rigt))) {
  row_i = death_WI_3races_sum_rigt[i,]
  age = row_i[1] # AgeGroup, i
  race = row_i[3] # RaceEthnicity, g
  upper = row_i[4] # Deaths_weighted
  lower = r_igt_lower_func(age = age, race = race)
  r_igt_i = as.numeric(upper) / lower
  death_WI_3races_sum_rigt$r_igt[i] = r_igt_i
}

death_WI_3races_sum_rigt  # row 76-90 are all ages 
```

$$
R_{g,t} = \sum_i  (r_{i,t,g} \times w_i)
$$

```{r}
R_gt_func = function(race, quarter){
  df = death_WI_3races_sum_rigt %>%
    filter(RaceEthnicity == race & Quarter == quarter)  # determine race & quarter
  sum_i = 0 # prepare for add up
  
  for (i in age_ranges) { # 5 age groups in total 
                          # age_ranges are characters, in the same form as those of death_??, and WIso index of w_i
    df_idx = which(df$AgeGroup == i) # just in case the order is different 
    
    if (length(df_idx) > 0){ # it is possible that our subset has severWI age groups missing, then df_idx = integer(0)
    aaa = df$r_igt[df_idx] * w_i[i]} # $(r_{i,t,g} \times w_i)$
    else {aaa = 0}
    
    sum_i = sum_i + aaa
  }
  
  ans = sum_i
  return(ans)
}
```

```{r}
length(unique(death_WI$Quarter))
unique(death_WI$Quarter)

R_gt = data.frame(racial_group = rep(c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic"), each = 5), time = rep(c("Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5"), 3))

R_gt = R_gt %>%
  mutate(R_gt_ = 0)

for (i in c(1:nrow(R_gt))) {
  race = R_gt$racial_group[i]
  quarter = R_gt$time[i]
  
  rrgti = R_gt_func(race = race, quarter = quarter)
  R_gt$R_gt_[i] = rrgti
}

R_gt
```


```{r}
ggplot(R_gt, aes(x = time, y = R_gt_))+
  geom_point()+
  facet_wrap(~ racial_group)+
  labs(x = NULL)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```